Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2020-05-26 doc/response.tex

\revtext{Lines 49,101: "...the intrinsic generation interval distribution..." While the
interval itself is not defined in Line 46 (in my opinion), its distribution is
clarified in Line 101 in relation to equation (2) under homogeneous mixing.
I hereby pose the following questions:}

\revtext{
1. Although $g(\tau) > 0$ satisfies all the properties of a probability density
function (p.d.f.) of a non-negative continues random variable, does that
automatically correspond to a specific random variable relevant to any of
the observable (or conceptually observable) "generation intervals"?
}

\revtext{
2. Is it the p.d.f. corresponding to the definition in Lines 42-43? Proof?
}

Our new definition of the generation interval and explanation in Section 2 (just above) address these questions to some extent. 
The definition in Lines 42-43 we had earlier does not correspond to any p.d.f.
We prefer not to go deeper into the mathematical details for this paper.

@@

We now show that the initial forward generation-interval distribution is equivalent to the intrinsic distribution.
We also point to the analytical intractability of the spatial effects on the realized generation-interval distributions (and therefore the initial forward distribution):

\newtext{Since spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles), we rely on simulations hereafter to vadliate our prediction.}

\revtext{However, I suggest the authors to compare the "individual-based
method" in the paragraph starting from Line 252 to Line 260 with the statistical methods for analyzing right-truncated survival data as mentioned
above, and explicitly write out the likelihood function (Line 258) to see if
they are the same or different.}

Thank you for pointing this out. 
We realized that we were implicitly combining two likelihoods, one of which is based on the non-homogeneous Poisson process and the other is based on truncated distributions (described in suggested references).
We now provide a more detailed explanation of the method with appropriate references.
We also allude to the fact that our proposed likelihood is a special case of the likelihood derived in Kalbfleisch and Lawless (1989).
The likelihood function is provided in the Methods section.

@@

We no longer use $\RR$ to avoid confusion. We now use $\Rini = \Ro S(0)$ to convey the idea that we consider the possibility that $S(0)$ may not be 1 but are not explicitly considering how $S$ changes through time.

\revtext{Is $S(t)$ a number or a proportion? In deterministic models, this matters
less as one can argue that all deterministic models are describing the
change of proportions (by re-scaling on population size, assuming it is
constant); but it matters a lot in stochastic models because the population
size is an essential parameter.}

We have tried to clarify these points. $S$ is a proportion, but we do attempt to address issues of individuality appropriately.

\swp{What do you mean we have tried to address issues of individuality? Can we just get rid of that phrase and say S is a proportion?}

@@

We had omitted a sentence previous --- we have now clarified this sentence as follows:

\newtext{In the main text, we present results based on exponentially distributed latent and infectious periods;
we show results based on more Erlang distributed latent periods ($n_E=2$), which better matches the incubation period distribution of Ebola virus disease, in Appendix.}

\revtext{
It is more convenient to use the Laplace transforms as a tool.
}

The reviewer's discussion of and suggestions about Laplace transformations are interesting, but we think beyond the scope of this paper. Our experience has been that Laplace transformations and moment generating functions are as likely to obscure the key concepts underlying disease dynamics as to clarify them.

@@

\revtext{How to assess if the gamma distribution is a good choice as a
model? Run some kind of goodness-of-fit tests?}

These are hard issues, but beyond the scope of this paper. 
Using exponentials is routine in many applications. 
Using gammas (or Erlangs) generalizes this, and is also commonly done. 
The gamma distribution has been previously tested for many diseases. 
In our case, Figure 5 demonstrates that the gamma approximation of the generation-interval distribution provides a reasonable estimate for the initial reproduction number from the exponential growth rate.
We now discuss additional evidence that gammas often provide reasonable estimates [our approximation framework] as well as potential caveats in assuming a shape of the generation-interval distribution in Discussion.

\newtext{We compare two methods for estimating the initial forward generation-interval distribution and assume that the initial forward generation-interval distribution follows a gamma distribution.
The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
we previously showed that summarizing the entire distribution with two moments (mean and variance) can be sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ for Ebola, rabies, and measles \citep{park2019practical}.
However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.4), even though the estimated gamma distribution may ``look'' indistinguishable from the true shape of the intrinsic generation-interval distribution (derived from the SEIR model).
These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estimate of \Rini (e.g., shorter mean generation interval leads to lower \Ro whereas narrower generation-interval distribution leads to higher \Ro) and, therefore, may have small effects on the overall estimate of $\Rini$ (Appendix A.4).}

@@

\revtext{
How exactly is $\RR$ (and, is it \Ro or some kind of time-dependent \RR) estimated in this paper: via the Euler-Lotka equation or using the approach in White and Pagano (2007, ref. [54] in the paper)?
}

We estimate $\Rini$ via Euler-Lotka. We have tried to make this clear throughout the main text.

\revtext{
How close is the "related approach" in White and Pagano
(2007, ref. [54] in the paper) related to the current paper? Did they use the
Euler-Lotka equation at all? Did they define and measure the generation intervals that were consistent with the generation-intervals discussed in
the current paper?
}

We had included this reference in earlier versions of this manuscript a few years ago, thinking that their approach is similar due to their use of Poisson likelihood.
We realized that their approach is quite different from the goal of our paper and have removed the sentence.
They do not use Euler-Lotka or rely on ``observed'' generation intervals;
instead they try to infer a generation-interval distribution from incidence reports.

@@

\revtext{
The authors use a logistic function with a scale parameter $r$,
the carrier capacity parameter $K$ and the initial condition $c_0$ to model the
cumulative incidence $c(t)$: Assuming, (theoretically), the infection events
are observable. It is inaccurate to claim that r is the exponential growth
rate
}

We have added the following justification to the text:

\newtext{
We estimate the initial exponential growth rate $r$ of an epidemic from daily incidence by modeling the cumulative incidence $c(t)$ with a logistic function \citep{ma2014estimating}:
\begin{equation}
c(t) = \frac{K}{1 + \left[(K/c_0)-1\right]e^{-rt}}.
\end{equation}
While the exponential growth rate (i.e., the rate of change in log incidence) of the logistic function changes throughout an epidemic,
we focus strictly on estimating the \emph{initial} exponential growth rate (when $t \to -\infty$).
The method of estimating the initial exponential growth rate by fitting a logistic curve has been previously validated against simulations of stochastic comparmental models \citep{ma2014estimating}.
}

\revtext{
The paper focuses on estimating the generating-interval distribution. There are
several generating-intervals and distributions mentioned in this paper, such as
$g(\tau)$; $b_t(\tau)$; $c_t(\tau)$; $g_{eff}(\tau)$; etc. To the end, it seems that the primary focus is the
intrinsic distribution $g(\tau)$; and relating this distribution with two other intrinsic
parameters: the reproduction number $\Ro$ and the Malthusian number $r$
}

@@

The intrinsic distribution is not our primary focus. 
As we show in Figure 4, the intrinsic distribution overesitmates $\Rini$ as it does not account for spatial effects.
We tried to make clear that our goal is the initial forward distribution.

\revtext{
The authors have well demonstrated different sources of sampling bias
arising from contact tracing data, more noticeably, (i) length-bias due
to right-truncation because of retrospective measurement together with
different infection times during the course of the outbreak; (ii) due to the
network structure (e.g. the degree distribution and the degree excess (Line
408). Both sources, the right-truncation and the egocentric point of view
(Line 174), tend to give shorter on average observed intervals. These are
correct and important.
}

\revtext{
Any notion of bias is in the context against some "truth". The authors
proposed, and demonstrated by simulation, ways to adjust for the bias in
order to estimate this "truth". However, the authors equate the "true"
distribution to the intrinsic distribution $g(\tau)$ as if "obvious". It is not
obvious and this needs to be proven.
The intrinsic distribution $g(\tau)$ is an equilibrium distribution that
involves the equilibrium distribution of the infectious period (in SEIS,
SEIR, SEIRS settings). It is a renewal process concept and is only
valid when the system is under the equilibrium condition. It is not
obvious how it is related to measurements of "generation intervals"
when the system is not at equilibrium.
}

We now discuss this point in much more detail (and have discussed these points above). Briefly: we are interested in the initial forward distribution because this provides the link between $r$ and \Rini. This approaches the intrinsic distribution in the case where mixing is homogeneous and the initial number infectious is small. We hope we have fixed all of the places where we failed to explain ourselves clearly.

### 2020-05-26 doc/network.tex
done???

At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
While this definition is widely used in the literature, it is not connected directly to a population-level \emph{distribution}.
There are important distinctions to be made when defining generation-interval distributions at the population-level. The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.
On the other hand, the \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.

@@

The population-level egocentric kernel is found by integrating the individual-level egocentric over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) \,\mathrm{d}a,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual has not been infected by the focal individual.
Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s)\,\mathrm{d}s}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair; for convenience, we will often omit ``population-level''.
Finally, the link between the initial exponential growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
\begin{equation}
\frac{1}{\Rego} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, $\Rego$ will be typically smaller than $\Rini$ estimated from the intrinsic distribution;
this generation-interval-based argument provides an alternative biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

@@

We compare two methods for estimating the initial forward generation-interval distribution and assume that the initial forward generation-interval distribution follows a gamma distribution.
The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ \citep{park2019practical}.
However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.4), even though the gamma distribution ``looks'' indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model).
These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estimate of \Rini (e.g., shorter mean generation interval leads to lower \Ro whereas narrower generation-interval distribution leads to higher \Ro) and, therefore, may have small effects on the overall estimate of \Rini (Appendix A.4).

@@

To study the effects on right-truncation on the observed generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
The SEIR model describes how disease spreads in a homogeneously mixing population; it assumes that infected individuals become infectious after a latent period.
We use a \seminar\ model, which extends the SEIR model to have multiple equivalent stages in the latent and infectious periods. This gives latent and infectious periods with Erlang distributions (gamma distributions with integer shape parameters, including the exponential distribution), which are often more realistic than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}: 
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \sum_{k=1}^{n_I} I_k\\
\frac{dE_1}{dt} &= \beta S \sum_{k=1}^{n_I} I_k - n_E \sigma E_1\\
\frac{dE_m}{dt} &= n_E \sigma (E_{m-1} - E_m) && \text{for } m = 2, 3, \dots, n_E\\
\frac{dI_1}{dt} &= n_E \sigma E_m - n_I \gamma I_1\\
\frac{dI_n}{dt} &= n_I \gamma (I_{n-1} - I_n) && \text{for } n = 2, 3, \dots, n_I\\
\end{aligned}
\end{equation}
where $S$ is the proportion of susceptible individuals, $E_m$ is the proportion of exposed individuals in the $m$-th compartment, $I_n$ is the proportion of infectious individuals in the $n$-th compartment, and $R$ is the proportion of recovered individuals.
Parameters of the model are specified as follows:  $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, $n_E$ is the number of latent compartments, $1/\gamma$ is the mean infectious period, and $n_I$ is the number of infectious compartments.
We scale the proportions of individuals in each compartment are scaled by the total population size $N$.
In the main text, we present results based on exponentially distributed latent and infectious periods;
we show results based on more Erlang distributed latent periods ($n_E=2$), which better matches the incubation period distribution of Ebola virus disease, in Appendix.

@@

The \textbf{individual-level method} models each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the truncation time ($\tsub{t}{truncate}$), with time-varying Poisson rate at time $t$ is equal to $\Lambda f_0(t - t_j)$, where $f_0(t)$ is the initial forward generation-interval distribution \citep{daley2007introduction}.
We use a gamma distribution (parameterized by its mean and shape) to model the initial forward generation-interval distribution.
Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{truncate}$ is equal to
\begin{equation}
\frac{\Lambda^{n_j} \exp \big(- \Lambda\, F_0(\tsub{t}{truncate} - t_j ;\theta) \big) \prod_{i=1}^{n_j} f_0(s_{i, j}; \theta)}{n_j!},
\end{equation}
where $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the initial forward generation-interval distribution $f_0$ (and the corresponding cumulative distribution function $F_0$).

The likelihood of the non-homogeneous Poisson process can be written as:
\begin{equation}\label{eq:nonhomogeneous}
\begin{aligned}
&\mathcal{L}(\Lambda, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{truncate})\\
&\quad=\prod_{j=1}^N \left(\Lambda^{n_j} \exp \big(- \Lambda \, F_0(\tsub{t}{truncate} - t_j ;\theta) \big) \prod_{i=1}^{n_j} f_0(s_{i, j}; \theta) \right),
\end{aligned}
\end{equation}
where $N$ is the total number of infected individuals.
Here, we estimate parameters $\Lambda$ and $\theta$ by maximum likelihood.
We then use the estimated distribution $f_0$ to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).
When the entire transmission process is known, we expect $\Lambda$ to match $\Rini$;
otherwise, $\Lambda$ will be sensitive to under-reporting of number of infections caused by each infected individual.
As we show in Appendix A.3, the estimates of $\Rini$ using the Euler-Lotka equation from $f_0$ remains unbiased even in the presence of under-reporting.

>>>>

At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
While this definition is widely used in the literature, it is not connected directly to a population-level \emph{distribution}.
There are important distinctions to be made when defining generation-interval distributions at the population-level. The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case \citep{champredon2015intrinsic}.
On the other hand, the \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.

@@

The population-level egocentric kernel is found by integrating the individual-level egocentric over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) \,\mathrm{d}a,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual has not been infected by the focal individual.
Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s)\,\mathrm{d}s}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair; for convenience, we will often omit ``population-level''.
Finally, the link between the initial exponential growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
\begin{equation}
\frac{1}{\Rego} = \int \hat{g}(\tau) \exp(-r \tau)\,\mathrm{d}\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, $\Rego$ will be typically smaller than $\Rini$ estimated from the intrinsic distribution;
this generation-interval-based argument provides an alternative biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

@@

We compare two methods for estimating the initial forward generation-interval distribution and assume that the initial forward generation-interval distribution follows a gamma distribution.
The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
we previously showed that summarizing the entire distribution with two moments (mean and variance) can be sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ for Ebola, rabies, and measles \citep{park2019practical}.
However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.4), even though the estimated gamma distribution may ``look'' indistinguishable from the true shape of the intrinsic generation-interval distribution (derived from the SEIR model).
These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estimate of \Rini (e.g., shorter mean generation interval leads to lower \Ro whereas narrower generation-interval distribution leads to higher \Ro) and, therefore, may have small effects on the overall estimate of $\Rini$ (Appendix A.4).

@@

To study the effects on right-truncation on the observed generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
The SEIR model describes how disease spreads in a homogeneously mixing population; it assumes that infected individuals become infectious after a latent period.
We use a \seminar\ model, which extends the SEIR model to have multiple equivalent stages in the latent and infectious periods. This gives latent and infectious periods with Erlang distributions (gamma distributions with integer shape parameters, including the exponential distribution), which are often more realistic than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}: 
\begin{equation}
\begin{aligned}
\frac{\mathrm{d}S}{\mathrm{d}t} &= - \beta S \sum_{k=1}^{n_I} I_k\\
\frac{\mathrm{d}E_1}{\mathrm{d}t} &= \beta S \sum_{k=1}^{n_I} I_k - n_E \sigma E_1\\
\frac{\mathrm{d}E_m}{\mathrm{d}t} &= n_E \sigma (E_{m-1} - E_m) && \text{for } m = 2, 3, \dots, n_E\\
\frac{\mathrm{d}I_1}{\mathrm{d}t} &= n_E \sigma E_m - n_I \gamma I_1\\
\frac{\mathrm{d}I_n}{\mathrm{d}t} &= n_I \gamma (I_{n-1} - I_n) && \text{for } n = 2, 3, \dots, n_I\\
\end{aligned}
\end{equation}
where $S$ is the proportion of susceptible individuals, $E_m$ is the proportion of exposed individuals in the $m$-th compartment, $I_n$ is the proportion of infectious individuals in the $n$-th compartment, and $R$ is the proportion of recovered individuals.
Parameters of the model are specified as follows:  $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, $n_E$ is the number of latent compartments, $1/\gamma$ is the mean infectious period, and $n_I$ is the number of infectious compartments.
We scale the proportions of individuals in each compartment are scaled by the total population size $N$.
In the main text, we present results based on exponentially distributed latent and infectious periods;
we show results based on more Erlang distributed latent periods ($n_E=2$), which better matches the incubation period distribution of Ebola virus disease, in Appendix.

@@

The \textbf{individual-level method} models each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the truncation time ($\tsub{t}{truncate}$), with time-varying Poisson rate at time $t$ is equal to $\Lambda f_0(t - t_j)$, where $f_0(t)$ is the initial forward generation-interval distribution \citep{daley2007introduction}.
We use a gamma distribution (parameterized by its mean and shape) to model the initial forward generation-interval distribution.
Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{truncate}$ is equal to:
\begin{equation}
\frac{\Lambda^{n_j} F_0(\tsub{t}{truncate} - t_j ;\theta)^{n_j} \exp \big(- \Lambda\, F_0(\tsub{t}{truncate} - t_j ;\theta) \big)}{n_j!},
\label{eq:nPoisson}
\end{equation}
where $\theta$ is a (vector) parameter of the initial forward generation-interval distribution $f_0$ (and the corresponding cumulative distribution function $F_0$).
On the other hand, the probability density that the realized generation-interval between infector $j$ and infectee $i$ is equal to $0 \leq s_{i,j} \leq \tsub{t}{truncate} - t_j$ can be expressed using a truncated distribution \citep{lagakos1988nonparametric, kalbfleisch1991regression}:
\begin{equation}
\frac{f_0(s_{i, j}; \theta)}{F_0(\tsub{t}{truncate} - t_j ;\theta)}.
\label{eq:truncdist}
\end{equation}
Therefore, the probability density that individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{truncate}$ with realized generation intervals $s_{i,j}$ for $i=1,\dots,n_j$ is a product of \eref{nPoisson} and \eref{truncdist}:
\begin{equation}
\frac{\Lambda^{n_j} \exp \big(- \Lambda\, F_0(\tsub{t}{truncate} - t_j ;\theta) \big) \prod_{i=1}^{n_j} f_0(s_{i, j}; \theta)}{n_j!}.
\end{equation}

The full likelihood of contact tracing data, which include aggregated generation intervals as well as information about who infected whom, until time $\tsub{t}{truncate}$ can be written as:
\begin{equation}\label{eq:nonhomogeneous}
\begin{aligned}
&\mathcal{L}(\Lambda, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{truncate})\\
&\quad=\prod_{j=1}^N \left(\frac{\Lambda^{n_j} \exp \big(- \Lambda\, F_0(\tsub{t}{truncate} - t_j ;\theta) \big) \prod_{i=1}^{n_j} f_0(s_{i, j}; \theta)}{n_j!} \right),
\end{aligned}
\end{equation}
where $N$ is the total number of infected individuals.
This likelihood is a special case of the likelihood suggested in \citep{kalbfleisch1989inference} which assumes that the second event (observation of infection) occurs simultaneously with the corresponding initiating event (infection of an individual).
Here, we estimate parameters $\Lambda$ and $\theta$ by maximum likelihood.
We then use the estimated distribution $f_0$ to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).
When the entire transmission process is known, we expect $\Lambda$ to match $\Rini$;
otherwise, $\Lambda$ will be sensitive to under-reporting of number of infections caused by each infected individual.
As we show in Appendix A.3, the estimates of $\Rini$ using the Euler-Lotka equation from $f_0$ remains unbiased even in the presence of under-reporting.
